---
title: ""
engine: knitr
---

------------------------------------------------------------------------

This step-by-step guide will walk you through using [**malariasimulation**](https://mrc-ide.github.io/malariasimulation/) in R to model malaria transmission and interventions and how to use the [**malariaverse**](https://mrc-ide.github.io/malariaverse/) to apply **malariasimulation** to a specific context. Let's dive right in!

------------------------------------------------------------------------

### Set up and installation

Firstly you'll need to get up and running with R and install the packages we will use.

```{r, eval=FALSE}
# Define where to look for package installations
options(
  repos = c(
    mrcide = 'https://mrc-ide.r-universe.dev',
    CRAN = 'https://cloud.r-project.org'
  )
)

# Install some packages
## The malaria model
install.packages("malariasimulation")
## Malaria model output processing help
install.packages("postie")
## For output wrangling
install.packages("dplyr")
## For plotting
#install.packages("ggplot2")
```

```{r}
# Load packages
library("malariasimulation")
library("postie")
library("dplyr")
library("ggplot2")

# Set a seed for reproducible stochastic outputs
set.seed(12345)
```

------------------------------------------------------------------------

### Starting Simple: Running the Malaria Model

To begin, we'll run a basic **malariasimulation** model simulation. This initial run will be generic — meaning it's not yet customised to represent a specific region, population or transmission setting. We'll also skip any interventions for now, so we're simply simulating transmission.

```{r}
# Get the default parameters
parameters <- get_parameters()
# Run a simple malariasimulation for 5 years
output <- run_simulation(
  timesteps = 365 * 5,
  parameters = parameters
)
```

We can take a look at the raw simulation output

```{r}
glimpse(output)
```

There are a lot of outputs for each timestep (day)!

------------------------------------------------------------------------

#### Analysing Outputs with `postie`

To get a sense of what's going on in our simulation, we can use the [**postie**](https://mrc-ide.github.io/postie/) package to process the results, ready for visualisation. This package will help you extract and format key metrics like prevalence, severe cases, and more.

```{r}
# Extract formatted prevalence from the raw output and add a run name
processed_output <- get_prevalence(
  x = output,
  baseline_year = 0
) |>
  mutate(run = "Baseline")
glimpse(processed_output)
```

Often, the best way to examine outputs is to plot them. Malaria prevalence indicates the "Proportion of a specified population with malaria infection at one time". We can plot *Plasmodium falciparum* parasite prevalence in 2-10 year (*Pf*PR~2-10~) olds over time.

```{r}
ggplot(
  data = processed_output,
  aes(x = time, y = lm_prevalence_2_10, colour = run)
) +
  geom_line() +
  scale_colour_manual(
    values = "#31035f", name = "Run"
  ) +
  xlab("Time (years)") +
  ylab(expression(italic(Pf) * Pr[2-10])) +
  ylim(0, 1) +
  theme_bw()
```

------------------------------------------------------------------------

#### Adjusting the intensity Transmission

In our model run above, we can see *Pf*PR~2-10~ hovering around 0.5. If we want, we can adjust the level of **baseline transmission** - that is, transmission intensity in the absence of interventions.

**malariasimulation** has lots of helper function to make it easier to modify the parameter list. One of these [`set_equilibrium()`](https://mrc-ide.github.io/malariasimulation/reference/set_equilibrium.html) allows us to adjust the baseline Entomological Inoculation Rate (EIR). The EIR is the "number of infective bites received per person in a given unit of time", in this case per year. Let's see what impact chnaging the EIR in our model has

```{r}
# Update the parameters with a new EIR
parameters <- parameters |>
  set_equilibrium(init_EIR = 2)

# Run a new simulation
output2 <- run_simulation(
  timesteps = 365 * 5,
  parameters = parameters
) 

# Extract formatted prevalence from the raw output and add a run name
processed_output2 <- get_prevalence(
  x = output2,
  baseline_year = 0
) |>
  mutate(run = "Modified EIR")

# Combined outputs from both simulations for plotting
plot_data <- bind_rows(
  processed_output,
  processed_output2
)

ggplot(
  data = plot_data,
  aes(x = time, y = lm_prevalence_2_10, col = run)
) +
  geom_line() +
    scale_colour_manual(
    values = c("#31035f", "#d32fbd"), name = "Run"
  ) +
  xlab("Time (years)") +
  ylab(expression(italic(Pf) * Pr[2-10])) +
  ylim(0, 1) +
  theme_bw()
```

Our `Baseline` run used the default EIR of 10. For our modified run, we can see the impact of lowering the EIR to 2 - *Pf*PR~2-10~ has fallen from around 50% to 20%.
